---
title: "475_final"
author: "Ruidong Chen"
date: '2022-12-10'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Name: Ruidong Chen  SID:488314**

**Name: Zhuoyun Chen  SID:499214**

**Name: Ziwei Liu     SID:489909**

**Name: Xinyang Feng  SID:490138**

For Monte Carlo methods, it is critical to uniformly sample from d-dimensional spheres and balls.

I will show some most frequent and useful methods to uniformly sample from the surface of a d-sphere and the interior of the d-ball.

**Definition of the d-spere and d-ball.**

A unit d-dimensional sphere is defined as follows:
$$
S^{d}=\{x\in R^{d+1}:|x|=1 \}
$$
And a unit d-dimensional ball is defined as follows:
$$
B^d=\{x\in R^d:|x|\leq1 \}
$$
Based on this definition, the perimeter of a circle is 1-sphere and the interior of the circle is 2-ball. The surface of a ball is 2-sphere and the interior of the ball is 3-ball. Also, the d signifies the degrees of freedom.

**Uniform Random Sampling**
For uniform random sampling, this means that all possible elements of S have an equal probability to be selected based on discrete probability. Besides, it also means the likelihood of an element falling in any sub-interval is discretely proportional to the length of the sub-interval based on continuous probabilities.

Also, a simple and common method to draw from a normal distribution is the Box-Muller algorithm.

To briefly understand why this works, consider the function
$$
f(u,v)=f(z)=(\frac{1}{\sqrt{2\pi}}e^{-0.5u^2})(\frac{1}{\sqrt{2\pi}}e^{-0.5v^2})
$$
After algebra simplifications
$$
f(u,v)=f(z)=\frac{1}{2\pi}e^{-0.5(u^2+v^2)}= \frac{1}{2\pi}e^{-0.5|z|^2}
$$
This shows that the probability distribution of z only depends on its magnitude and not any other direction. Thus it must be symmetric and must correspond to a uniform distribution on the circle.

This argument can be generalized to any d.

**1:Rejection Method**

This is probably the most intuitive method and is fast for 2-ball.

However, it will get a bad rap because when we apply this method to high dimensions, it can become very inefficient.

pseudo code:

1: Select x,y ~ U(-1,1)

2: If (x^2+y^2>1) then reject and go to step 1.

3: Return (x,y)

**2:Polar + Radial CDF**

For balls of any dimension d, a very common method of uniformly picking form a d-ball is to first select a random directional unit-vector from (d-1)-sphere and then multiply this vector by a scaler radial multiplicative factor.

The crucial aspect of this algorithm is the multiplicative factor square root function. This is because area which corresponds to the CDF grows according to r square. Therefore, we need to apply the inverse of this.

```{r}
set.seed(1)
u=runif(10000)
v=runif(10000)
r=u^0.5
theta=2*pi*v
x=r*cos(theta)
y=r*sin(theta)
colors <- c('#56B4E9')
plot(x,y,col=colors,pch = '.')

```

**3:Concentric Map**

This variance reduction method is motivated by the idea that ideal mappings from square-based random variate uv-space to the coordinate xy-space should be area-preserving, two-way continuous and with low distortion.

```{r}
set.seed(1)
u=runif(10000)
v=runif(10000)
if (u==0 && v==0) return(0,0)
theta=0
r=1
a=2*u-1
b=2*v-1
r=numeric(10000)
theta=numeric(10000)

for (i in 1:10000){
  if (a[i]*a[i]>b[i]*b[i]){
    r[i]=a[i]
    theta[i]=pi/4*b[i]/a[i]
  }else{
    r[i]=b[i]
    theta[i]=pi/2-pi/4*a[i]/b[i]
  }
}

x=r*cos(theta)
y=r*sin(theta)

colors <- c('#56B4E9')
plot(x,y,col=colors,pch = '.')

```
**4:Exponential Distribution**

Another method that is valid for balls of any dimension is to draw d random variates from the normal distribution and a single random variable from exponential distribution usually with lambda=0.5.

Also, a simple and common method to draw from an exponential distribution with parameter lambda is via the the negative logarithm of uniform random variate divided by lambda.

```{r}
u=rnorm(10000)
v=rnorm(10000)
e=rexp(10000,0.5)
denom=(u*u+v*v+e)^0.5
x=u/denom
y=v/denom
colors <- c('#56B4E9')
plot(x,y,col=colors,pch = '.')
```

**5:Dropped Coordinates**

A new method first introduced by Harman and Lacko in 2010 and then proven by Voelker in 2017 that is not famous is that if (x1,x2,x3,x4) is a random vector uniformly distributed on the 3-sphere, the random vector (x1,x2) is uniformly distributed in the 2-ball.

```{r}
a=rnorm(10000)
b=rnorm(10000)
c=rnorm(10000)
d=rnorm(10000)
norm=(a*a+b*b+c*c+d*d)^0.5
x=c/norm
y=d/norm
colors <- c('#56B4E9')
plot(x,y,col=colors,pch = '.')
```

**Uniformly sampling a 2-sphere which is the surface of a sphere**

We can uniformly sample points on the surface of a sphere.

**polar**

This is the natural generalization.
```{r}
library(scatterplot3d)
u=runif(10000)
v=runif(10000)
theta=2*pi*u
phi=acos(2*v-1)
x=sin(theta)*cos(phi)
y=sin(theta)*sin(phi)
z=cos(theta)
colors <- c('#56B4E9')
scatterplot3d(x,y,z,color=colors,pch = '.')
```
We can also use u=cosine theta which does not use inverse trig functions as follows:
```{r}
library(scatterplot3d)
u=2*runif(10000)-1
phi=2*pi*runif(10000)
z=u
x=cos(phi)*(1-z^2)^0.5
y=sin(phi)*(1-z^2)^0.5
colors <- c('#56B4E9')
scatterplot3d(x,y,z,color=colors,pch = '.')
```

**Muller method**
```{r}
library(scatterplot3d)
a=rnorm(10000)
b=rnorm(10000)
c=rnorm(10000)
norm=(a*a+b*b+c*c)^0.5
x=a/norm
y=b/norm
z=c/norm
colors <- c('#56B4E9')
scatterplot3d(x,y,z,color=colors,pch = '.')
```

**Uniformly sampling a 3-ball which is interior of a sphere**

We will now uniformly sample points inside a sphere.

**Muller method for 3-ball**

```{r}
library(scatterplot3d)
a=rnorm(10000)
b=rnorm(10000)
c=rnorm(10000)
r=runif(10000)^(1/3)
norm=(a*a+b*b+c*c)^0.5
x=r*a/norm
y=r*b/norm
z=r*c/norm
colors <- c("#E69F00")
scatterplot3d(x,y,z,color=colors,pch = '.')
```

**polar method for 3-ball**

```{r}
library(scatterplot3d)
u=2*runif(10000)-1
phi=2*pi*runif(10000)
r=runif(10000)^(1/3)
x=r*cos(phi)*(1-u^2)^0.5
y=r*sin(phi)*(1-u^2)^0.5
z=r*u
colors <- c("#E69F00")
scatterplot3d(x,y,z,color=colors,pch = '.')
```
**Exponential Distribution for 3-ball**

This is a similar method as 2-sphere.
```{r}
library(scatterplot3d)
a=rnorm(10000)
b=rnorm(10000)
c=rnorm(10000)
e=rexp(10000,rate=0.5)
denom=(e+a*a+b*b+c*c)^0.5
x=a/denom
y=b/denom
z=c/denom
colors <- c("#E69F00")
scatterplot3d(x,y,z,color=colors,pch = '.')
```

**Dropped coordinates for 3-ball**

Also, we have random vectors (a,b,c,d,e) uniformly distributed on the 4-sphere, the random vector (c,d,e) is uniformly distributed in the 3-ball and we dropped the first two coordinates.
```{r}
library(scatterplot3d)
a=rnorm(10000)
b=rnorm(10000)
c=rnorm(10000)
d=rnorm(10000)
e=rnorm(10000)
norm=(a*a+b*b+c*c+d*d+e*e)^0.5
x=c/norm
y=d/norm
z=e/norm
colors <- c("#E69F00")
scatterplot3d(x,y,z,color=colors,pch = '.')
```

**Other method**
```{r}
#install.packages("scatterplot3d")
```

```{r}
#r2 space
#multivariate approach
library(scatterplot3d)
n=2000
k=2
x=matrix(rnorm(n*k),ncol=k)
y=x/sqrt(rowSums(x*x))
colors <- c('#56B4E9')
plot(y[,1],y[,2],col=colors,pch = '.')
```
```{r}
# r3 space
#multivariate approach
n=10000
k=3
x=matrix(rnorm(n*k),ncol = k)
y=x/sqrt(rowSums(x*x))
colors <- c('#56B4E9')
scatterplot3d(y[,1],y[,2],y[,3],color=colors,pch = '.')
```

```{r}
#r2 space
#random numbers in a square
n=2000
k=2
x=matrix(runif(k*ceiling(n*4/pi),-1,1),ncol = k)
x=x[x[,1]^2+x[,2]^2<=1,]
y=x/sqrt(rowSums(x*x))
colors <- c('#56B4E9')
plot(y[,1],y[,2],col=colors,pch = '.')
dim(y)
```
```{r}
#r3 space
#random numbers in a cube
n=10000
k=3
x=matrix(runif(k*ceiling(n*6/pi),-1,1),ncol = k)
x=x[x[,1]^2+x[,2]^2+x[,3]^2<=1,]
y=x/sqrt(rowSums(x*x))
colors <- c('#56B4E9')
scatterplot3d(y[,1],y[,2],y[,3],color=colors,pch = '.')
dim(y)
```

**Generalization**: How to generalize to n-dimensional sphere or ball sampling? Using Muller Method.We can do it by random numbers in a cube method, but its efficiency drops quickly after d=3.

Start from sphere sampling.

**Muller Method**
```{r}
d_dimension_sphere = function(D) {
    samples = sapply(1:D, function(d) {
        rnorm(n=10000)
    })
    radii = apply(samples, 1, function(s) {
        sqrt(sum(s ^ 2))
    })
    samples = samples/radii
samples
}
```

```{r}
points.sphere = d_dimension_sphere(D=3)
x = points.sphere[, 1]
y = points.sphere[, 2]
z = points.sphere[, 3]
colors <- c('#56B4E9')
scatterplot3d(x,y,z,color=colors,pch = '.')
```
Ball sampling:

**Muller Method**
Similar to the 3-ball Muller, it just extend to higher dimension.
```{r}
d_dimension_ball = function(D) {
    samples = sapply(1:D, function(d) {
        rnorm(n=10000)
    })
    radii = apply(samples, 1, function(s) {
        sqrt(sum(s ^ 2))
    })
    new.radii = runif(10000) / radii
    samples = samples * new.radii
samples
}
```

```{r}
points.sphere = d_dimension_ball(D=3)
x = points.sphere[, 1]
y = points.sphere[, 2]
z = points.sphere[, 3]
colors <- c('#E69F00')
scatterplot3d(x,y,z,color=colors,pch = '.')
```

**Extension**: How to sample from unit hemisphere surface? (Given one of the coordinates is positive)

**Cosine-weighted hemisphere**
This is the natural generalization. Main difference from the sphere is the interval of theta to make sure theta range from (-pi/2,pi/2). In such a case, cos(theta)>0.
```{r}
#r-3 space
u=runif(10000)
v=runif(10000)
theta=acos(sqrt(u))
phi=2*pi*v

x=sin(theta)*cos(phi)
y=sin(theta)*sin(phi)
z=cos(theta)
colors <- c('#56B4E9')
scatterplot3d(x,y,z,color=colors,pch = '.')
```
We can also use u=cosine theta which does not use inverse trig functions as follows:
```{r}
u=runif(10000)
v=runif(10000)
a = (1 - 2*u)^2
b = sqrt(1 - a*a)
phi = 2*pi*v
x = b*cos(phi)
y = b*sin(phi)
z = a
colors <- c('#56B4E9')
scatterplot3d(x,y,z,color=colors,pch = '.')
```



